How is experience transformed into memory?

### 2018-09-03 paper/main.tex
corrected HeusEtal18 references

How might one go about formally characterizing the gist of an experience, or whether that gist has been recovered by the rememberer?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep[e.g., ][]{LernEtal11}.  Therefore the timecourse of how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Our memory systems can then leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we are in is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.  Although our experiences most often change gradually, they also occasionally change suddenly~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) during an experience help to discretize our experiences into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within event) and transient (across event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distill those experiences down to their essence-- i.e., their gists.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.

@@

The dynamic content of the video and participants' recalls is quantified in the corresponding topic proportion matrices.  However, it is difficult to gain deep insights into that content solely by examining the topic proportion matrices (e.g., Figs.~\ref{fig:model}A, D) or the corresponding correlation matrices (Figs.~\ref{fig:model}B, E, \corrmats).  To visualize the time-varying high-dimensional content in a more intuitive way~\citep{HeusEtal18} we projected the topic proportion matrices onto a two-dimensional space using Uniform Manifold Approximation and Projection~\citep[UMAP; ][]{McInHeal18}.  In this lower-dimensional space, each point represents a single video or recall event, and the distances between the points reflect the distances between the events' associated topic vectors (Fig.~\ref{fig:trajectory}).

@@

\subsection*{Modeling the dynamic content of the video and recall transcripts}
\subsubsection*{Topic modeling}
The input to the topic model we trained to characterize the dynamic content of the video comprised hand-generated annotations of each of 1000 scenes spanning the video clip~\citep[generated by][]{ChenEtal17}.  The features included: narrative details (a sentence or two describing what happened in that scene); whether the scene took place indoors or outdoors; names of any characters that appeared in the scene; name(s) of characters in camera focus; name(s) of characters who were speaking in the scene; the location (in the story) that the scene took place; camera angle (close up, medium, long, top, tracking, over the shoulder, etc.); whether music was playing in the scene or not; and a transcription of any on-screen text.  We concatenated the text for all of these features within each segment, creating a ``bag of words'' describing each scene.  We then re-organized the text descriptions into overlapping sliding windows spanning 50 scenes each. In other words, the first text sample comprised the combined text from the first 50 scenes (i.e., 1--50), the second comprised the text from scenes 2--51, and so on.  We trained our model using these overlapping text samples with \texttt{scikit-learn}~\citep[version 0.19.1; ][]{PedrEtal11}, called from our high-dimensional visualization and text analysis software, \texttt{HyperTools}~\citep{HeusEtal18}.  Specifically, we use the \texttt{CountVectorizer} class to transform the text from each scene into a vector of word counts (using the union of all words across all scenes as the ``vocabulary,'' excluding English stop words); this yields a number-of-scenes by number-of-words \textit{word count} matrix.  We then use the \texttt{LatentDirichletAllocation} class (topics=100, method=`batch') to fit a topic model~\citep{BleiEtal03} to the word count matrix, yielding a number-of-scenes (1000) by number-of-topics (100) \textit{topic proportions} matrix.  The topic proportions matrix describes which mix of topics (latent themes) is present in each scene.  Next, we transformed the topic proportions matrix to match the 1976 fMRI volume acquisition times.  For each fMRI volume, we took the topic proportions from whatever scene was displayed for most of that volume's 1500~ms acquisition time.  This yielded a new number-of-TRs (1976) by number-of-topics (100) topic proportions matrix.

@@

\subsubsection*{Visualizing the video and recall topic trajectories}
We used the UMAP algorithm~\citep{McInHeal18} to project the 100-dimensional topic space onto a two-dimensional space for visualization (Figs.~\ref{fig:trajectory}, \ref{fig:topics}).  To ensure that all of the trajectories were projected onto the \textit{same} lower dimensional space, we computed the low-dimensional embedding on a ``stacked'' matrix created by vertically concatenating the events-by-topics topic proportions matrices for the video and all 17 participants' recalls.  We then divided the rows of the result (a total-number-of-events by two matrix) back into separate matrices for the video topic trajectory and the trajectories for each participant's recalls (Fig.~\ref{fig:trajectory}).  This general approach for discovering a shared low-dimensional embedding for a collections of high-dimensional observations follows \cite{HeusEtal18}.

>>>>

How might one go about formally characterizing the gist of an experience, or whether that gist has been recovered by the rememberer?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep[e.g., ][]{LernEtal11}.  Therefore the timecourse of how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Our memory systems can then leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we are in is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.  Although our experiences most often change gradually, they also occasionally change suddenly~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) during an experience help to discretize our experiences into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within event) and transient (across event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distill those experiences down to their essence-- i.e., their gists.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.

@@

The dynamic content of the video and participants' recalls is quantified in the corresponding topic proportion matrices.  However, it is difficult to gain deep insights into that content solely by examining the topic proportion matrices (e.g., Figs.~\ref{fig:model}A, D) or the corresponding correlation matrices (Figs.~\ref{fig:model}B, E, \corrmats).  To visualize the time-varying high-dimensional content in a more intuitive way~\citep{HeusEtal18a} we projected the topic proportion matrices onto a two-dimensional space using Uniform Manifold Approximation and Projection~\citep[UMAP; ][]{McInHeal18}.  In this lower-dimensional space, each point represents a single video or recall event, and the distances between the points reflect the distances between the events' associated topic vectors (Fig.~\ref{fig:trajectory}).

@@

\subsection*{Modeling the dynamic content of the video and recall transcripts}
\subsubsection*{Topic modeling}
The input to the topic model we trained to characterize the dynamic content of the video comprised hand-generated annotations of each of 1000 scenes spanning the video clip~\citep[generated by][]{ChenEtal17}.  The features included: narrative details (a sentence or two describing what happened in that scene); whether the scene took place indoors or outdoors; names of any characters that appeared in the scene; name(s) of characters in camera focus; name(s) of characters who were speaking in the scene; the location (in the story) that the scene took place; camera angle (close up, medium, long, top, tracking, over the shoulder, etc.); whether music was playing in the scene or not; and a transcription of any on-screen text.  We concatenated the text for all of these features within each segment, creating a ``bag of words'' describing each scene.  We then re-organized the text descriptions into overlapping sliding windows spanning 50 scenes each. In other words, the first text sample comprised the combined text from the first 50 scenes (i.e., 1--50), the second comprised the text from scenes 2--51, and so on.  We trained our model using these overlapping text samples with \texttt{scikit-learn}~\citep[version 0.19.1; ][]{PedrEtal11}, called from our high-dimensional visualization and text analysis software, \texttt{HyperTools}~\citep{HeusEtal18a}.  Specifically, we use the \texttt{CountVectorizer} class to transform the text from each scene into a vector of word counts (using the union of all words across all scenes as the ``vocabulary,'' excluding English stop words); this yields a number-of-scenes by number-of-words \textit{word count} matrix.  We then use the \texttt{LatentDirichletAllocation} class (topics=100, method=`batch') to fit a topic model~\citep{BleiEtal03} to the word count matrix, yielding a number-of-scenes (1000) by number-of-topics (100) \textit{topic proportions} matrix.  The topic proportions matrix describes which mix of topics (latent themes) is present in each scene.  Next, we transformed the topic proportions matrix to match the 1976 fMRI volume acquisition times.  For each fMRI volume, we took the topic proportions from whatever scene was displayed for most of that volume's 1500~ms acquisition time.  This yielded a new number-of-TRs (1976) by number-of-topics (100) topic proportions matrix.

@@

\subsubsection*{Visualizing the video and recall topic trajectories}
We used the UMAP algorithm~\citep{McInHeal18} to project the 100-dimensional topic space onto a two-dimensional space for visualization (Figs.~\ref{fig:trajectory}, \ref{fig:topics}).  To ensure that all of the trajectories were projected onto the \textit{same} lower dimensional space, we computed the low-dimensional embedding on a ``stacked'' matrix created by vertically concatenating the events-by-topics topic proportions matrices for the video and all 17 participants' recalls.  We then divided the rows of the result (a total-number-of-events by two matrix) back into separate matrices for the video topic trajectory and the trajectories for each participant's recalls (Fig.~\ref{fig:trajectory}).  This general approach for discovering a shared low-dimensional embedding for a collections of high-dimensional observations follows \cite{HeusEtal18a}.
